---
title: "R Basics"
author: "Jonas, Pietro & Hauke"
format: 
  revealjs:
    theme: [simple, custom.scss]
    logo: img/logo.png
    footer: "[Institute of Behavioral Science and Technology](https://www.ibt.unisg.ch/)"
    preview-links: true
---
# Welcome to Session 2

```{r, include =FALSE}
        library(tidyverse)
```


## Recap Session 1
- Setting up R
- Get to know the UI
- Objects, classes, operations
- Read in data
- Hauke: what else?

## Any experienced difficulties? Anything we should repeat?


## Overview Session 2
1. Packages to the rescue
2. Read in and export,
3. Explore,
4. Manipulate, and 
5. Analyze data 

## Packages in R {.smaller}
Beyond `base` R's core functions, packages provide additional functionality for (almost) [all of your problems](https://support.rstudio.com/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages).


Some of the most widely-used packages are included in the [`tidyverse`](https://www.tidyverse.org/packages/#core-tidyverse) collection
`install.packages("tidyverse")` and `library(tidyverse)` brings you most of the functions you will need (at this point):

For example:

- manipulate data: `dplyr`
- tidy data: `tidyr`
- visualize data: `ggplot2`

There are many more packages [available](https://support.rstudio.com/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages)

::: aside
Info on any package can be found in its CRAN documentation.
:::

## Practical Advice {.smaller}
::: callout-note
## Pitfall
Failing to load a package is a common `error` source - one you can easily avoid by loading everything you need right at the start.
:::

This is how the very beginning of your code could look like:
```{r, include = T, echo=T}
#install.packages("ggplot2") # only needed for installing once
#install.packages("tidyverse") # only needed for installing once
library(ggplot2) # load package
library(tidyverse) # load package
```
::: aside
You might also directly specify a package (without loading it globally) by using the following syntax: `ggplot2::ggplot()`
:::

## Recap: Opening our code we need to have some things in place
Set your working directory, load your packages:
```{r, include = T, echo=T}
# setwd("/Users/haukeroggenkamp/Documents/coursework/r-basics") #set your working directory
#install.packages("ggplot2") # only needed for installing once
#install.packages("tidyverse") # only needed for installing once
library(ggplot2) # load package
library(tidyverse) # load package
```

## What's next: We need some data to work with!
We will learn to:

- read in and store, 
- explore,
- manipulate,
- analyze and 
- visualize data. <br/>


## Reading in and storing data {.smaller}
Data can come in various file formats, e.g., `.csv`, `.xlsx`, `.sav` etc. We can read all these files into our `environment`. 

For that, we always specify the location path to the local data source (and we directly assign our data to an object called `data`)

:::{.columns}
::: {.column width="50%"}
Reading in
```{r, echo=TRUE}
data <- read.table("/Users/haukeroggenkamp/Documents/coursework/r-basics/student_experiment.csv", sep = ";", header = T) # you need to adapt this to your working directory file path
```
::: 

::: {.column width="50%"}
Saving
```{r, echo=T}
write.table(data, "/Users/haukeroggenkamp/Documents/coursework/r-basics/student_experiment_processed.csv") # writes the final dataframe as a .csv file into your specified directory
```
:::
:::

::: callout-note
## Good to know
We recommend text files (`.csv` or `.txt`), however, there are packages for reading in other file formats (`haven`, `foreign` or `rio`).
:::

## Exploring data I/II {.smaller}
You should know how your data is structured before processing it and R has some neat functions to do that:

- `head(data)` shows the first few rows of your data
- `names(data)` shows the column or variable names of your data
- `view(data)` shows the entire data in a new window

```{r, echo =T, eval=F}
head(data)
names(data)
```

```{r, echo=FALSE, eval=T}
head <- head(data)
head[1:5, 1:5]
names <- names(data)
names[1:10]
```



## Exploring data II/II {.smaller}
To get a feeling for your data, you can investigate summary statistics:

::: {.columns}
::: {.column width="50%"}
```{r echo=T}
summary(data$rent) 
```
:::

::: {.column width="50%"}
```{r echo=T}
rstatix::get_summary_stats(data, rent, type="five_number") 
```
:::
:::

We can calculate the mean rent of this course individually:
```{r echo = T}
mean(data$rent)
```
How about the course's gender distribution?
```{r, echo=TRUE}
table(data$gender) # 1 = male, 2 = female
```
Or the mean shoe size by gender?
```{r, echo=T}
stats::aggregate(shoesize ~ gender, data, FUN = mean)
```

## Manipulate your data {.smaller}
Some important functions from `dplyr`:

- `select` Choose which columns to include.
- `filter` Filter which rows to include.
- `arrange` Sort the data, by size for continuous variables, by date, or alphabetically.
- `group_by` Group the data by a categorical variable.
- `n()` Count the number of records. Here there isn’t a variable in the brackets of the function, because the number of records applies to all variables.
- `mutate` Create new column(s) in the data, or change existing column(s).
- `rename` Rename column(s).
- `bind_rows` Append one data data frame to another, combining data from columns with the same name.

## Manipulate: `select` and `filter`{.smaller}

```{r echo = T, eval=FALSE}
data %>% 
  select(gender, rent) # show just the specified columns
```
```{r}
t <- data %>% 
  select(gender, rent) # show just the columns gender and rent
t[1:3,]
```


```{r echo = T}
data %>%
  filter(gender==1 & shoesize ==44) %>% select(gender, shoesize) # show just males with a shoesize of 44
```
::: callout-note
## About the `%>%`(pipe) operator
`%>%` makes code easier to write and read. It works similar to a + sign and is an integral part of the tidyverse syntax
:::


## Manipulate: `arrange` and `n`{.smaller}
```{r echo = T}
data %>% 
     filter(gender == 1 & rent <= 900) %>% # only males with rent below 900CHF
     arrange(desc(rent)) %>% # sort by descending rent 
        select(gender, rent) # show only the columns gender and rent
```

```{r echo = T}
data %>%
  filter(gender == 2) %>% # only female
  summarise(
    n = n()) # number of instances written in new column with name n
```

::: callout-note
## `summarise()`
`summarise()` reduces multiple values down to a single summary.
:::

## Manipulate: `group_by`{.smaller}
```{r echo = T}
data %>%
  group_by(gender) %>% # group by gender
  summarise(
    n = n(), # counte the instances and write in column named n
    avg_height = mean(height, na.rm = TRUE),
    avg_rent = mean (rent, na.rm = TRUE)) # calculate the mean height and write in avg_height column
```

::: callout-note
## Removing `NAs`
`na.rm` allows you to exclude instances of missing value. This helps to prevent `error` messages. But check why is that data missing!
:::

## Manipulate: `mutate` and `rename`{.smaller}

```{r echo = T, eval=F}
data %>% 
  mutate(gender, sq_height = height ^ 2) %>% # adds variable that is a function of existing ones
  select(gender, sq_height) # ashow only the columns gender and the new variable
```

```{r}
t <- data %>% 
   mutate(gender, sq_height = height ^ 2) %>% # adds variable that is a function of existing ones
  select(gender, sq_height)
t[1:5,]
```

```{r echo = T, eval = F}
data %>% 
  rename(
    cost_for_living = rent, # renames a column (new name = old name)
    )
```
```{r}
t <- data %>% 
  rename(
    cost_for_living = rent,
    )
t[1:5,1:5]
```

## Put it all together{.smaller}
```{r echo = T}
data %>%  # Start with the data 
               filter(r_experience <= 6) %>% # Oonly those with R experience below 6
               group_by(gender) %>%    # Group by gender
               summarise( 
                        age.mean = mean(age),      # Define first summary...
                        height.mean = median(height), # you get the idea...
                        n = n()) %>% # How many are in each group?
        select(age.mean, height.mean, n)
```

::: callout-note
## Help
It can sometimes be hard to understand what a command is doing with the data. [Tidy data tutor](https://tidydatatutor.com/vis.html) visualizes what is happening to your data in every code step which is extremely helpful.
:::


## Advanced: Join two dataframes{.smaller}

- `inner_join`: To keep only rows that match from the data frames, specify the argument `all=FALSE`. To keep all rows from both data frames, specify `all=TRUE`.
- `left_join`: To include all the rows of your data frame x and only those from y that match.

Example: Calculate year born of members of this class:
```{r, include = F}
#library(clipr) # load a clipboard copy tool to export right panel age table from #https://irp-cdn.multiscreensite.com/3438c060/files/uploaded/Age%20Calculator%202021_rev.pdf
#age_table <- read_clip_tbl(sep="\t", header = F
#) # load data from clipboard
#age_table <- age_table %>% separate(V1, c('year_born', '2021_before','2021_after', '2022_before', #'2022_after', '2023_before', '2023_after' )) # process data for join
#age_table <- as.data.frame(sapply(age_table, as.numeric)) # make columns numeric to join and same object #class
#write_csv(age_table, "~/TechX Dropbox/IBT/Teaching/R Basics/age_table.csv")
```


```{r echo=T}
age_table <- read.csv("age_table.csv") # read in age table from directory
data_new <- left_join(data, select(age_table, c(X2022_after, year_born)), by = c("age" = "X2022_after")) # perform join to get the birthyear
data_new %>% select(age, year_born) %>% filter(age<24) # show the results
```

::: aside
[Here](https://paldhous.github.io/ucb/2018/dataviz/week7.html) is a great tutorial on applying these functions.
:::


## Advanced: Long to wide dataformat {.smaller}

::: {.columns}
::: {.column width="50%"}
`data_long` to `data_wide`
```{r echo = T}
ncol(data) # number of columns in long format
data_wide <- spread(data, gender, height) # gather to wide format
ncol(data_wide) # number of columns in wide format
```

:::

::: {.column width="50%"}
The arguments to spread():

- data: Data object
- key: Name of column containing the new column names
- value: Name of column containing values

::: callout-note
## Reverse Way
The reverse way from wide to long format via `gather`
:::
:::
:::

## Analyze your data: Statistical Analysis with R {.smaller}

You (R) can do a lot:

-   _t-Test_
-   _ANOVA_
-   _Regression_
-   Structural Equation Models
-   Multilevel Models
-   Machine Learning Models
-   Bayesian Statistics
-   ...


## Stats 1: t-Test {.smaller}

Recap: Mean comparisons (>= 2 groups)

```{r echo=T}
t.test(formula = rent ~ gender, data = data, alternative = 'two.sided') # base R
```

```{r echo=T}
rstatix::t_test(data = data, formula = rent~gender, alternative = "two.sided") # the same works with other packages, here `rstatix`
```
Notice the difference in the code?

::: callout-note
## Caution
There are similar functions `t.test` and `t_test` from *different* packages using *different* syntax
:::

## Stats 2: Analysis of Variance {.smaller}

Recap: Mean comparisons (> 2 groups)

Is there a difference in XYZ?
```{r echo=T}
data$course <- as.factor(data$education) # this ensures our factor to be of the right class
summary(aov(formula = rent~education, data = data))
```

::: callout-note
## Caution
The underlying grouping variable needs to be `as.factor`. Always check first!
:::

```{r echo=T}
anova(lm(formula = rent~education, data = data)) # Same analysis possible by running a linear model and wrapping the `anova` function for omnibus test
```

## Stats 3: Regression {.smaller}

Recap: Relationship between two (or more) variables

Is there a relationship between height and shoesize?
```{r echo=T}
summary(lm(formula = shoesize~height, data = data)) # runs a linear model and provides statistical summary
```

## Live-saving advice: How to get help when stuck? {.smaller}

- `?lm()` will immediately provide you with Help in the panel
- [CRAN](https://cran.r-project.org/) documentations and Cheat Sheets are a comprehensive source of support (hard to read at the beginning) 
- Google is your friend: Chance is high that someone else had a very similar question to yours on [stackoverflow.com](https://stackoverflow.com/questions/tagged/r); try different search terms and exact language

:::aside
R studio provides a long list of [cheat sheets](https://www.rstudio.com/resources/cheatsheets/)
:::

## Jump in: Time for exercising
Advanced exercises: Calculate the shoe soze converted in the US system

- `shoesizes` dataframe
